Grove: A Convergent Collaborative Structure-Editor Calculus

### 2020-08-10 grove-paper.tex
Work on Single-User Edits

@@

TODO: somewhere in here, maybe multiple places, we want to talk about typing the intermediate states that arise

TODO: annotate vertexes and edges with ids (gray superscript?)

Single User Intuition:
 - manipulating trees,
 - holes

Multi User Intuition:
 - vertexes have unique ids
 - edits create or delete edges in the tree,
 - vertexes are automatically created when an edge goes to it

Deeper intuition:
 - edges have unique ids
 - once deleted an edge can never be recreated (though, a new edge (with a new id) can be created in the same spot as the original)

For presentational simplicity we omit param\_type in all lambdas

\subsection{Structure Editing}

We present this at two levels.
One is the underlying graph model, which
involves explicit identifiers, graph edges, indexes, and add and delete
operations.

The other is the interface a user might see,
which omits these details unless needed.

Throughout these examples the graph model is presented
as a graph with.

\subsubsection{Trees and Holes}
\label{sub:Trees and Holes}

TODO: explain holes.

\begin{figure}
\centering 
\includegraphics[scale=0.75]{graphviz/simple-tree-with-hole/1.dot}
\caption{A simple tree with a hole}
\label{fig:A simple tree with a hole}
\end{figure}
NEEDS UI EXAMPLE

Consider the tree in Figure~\ref{fig:A simple tree with a hole},
which represents the code \texttt{2 + \hole}.

Note that this tree contains three vertexes.
The first is depicted with a simple dot and is the root vertex.
The other two are labeled \texttt{plus} and \texttt{num(2)}
and represent the plus operator and the numeric literal 2, respectively.

Each vertex has a unique identifier and a list of positions
for child vertexes.
In this figure, the \texttt{plus} and \texttt{num(2)} vertexes have
identifiers 3 and 6 respectively,
the \texttt{plus} vertex has positions for its left and right children
and the \texttt{num(2)} vertex has no positions for any children.
The root vertex has the distinguished identifier 0
and has one position for its children.

Also each edge has a unique identifier. In Figure~\ref{fig:A simple tree with a hole},
these edges have idenifiers 4 and 7.  These are used
when adding and deleting edges as discussed in the
next subsubsection.

In this paper we represent variables and numeric constants
as leaves and represent their names or values as
arguments to those constructors (e.g., \texttt{2} in \texttt{num(2)}).
(In TODO:REF, we discuss how to allow edits within these.)

TODO: web page Svg ?

Declarative sentences

Outline refinement

The edge coming out of it to the TODO:word of the expression.
We call this the \emph{rooted} vertex.

\subsubsection{Delete and Add}

@@

@@

@@

\subsection{Collaborative Structure Editing}

@@

\begin{figure}
\centering
lam x. app(\#1, \hole)
\caption{Single-User Cycles}
\label{fig:Single-User Cycles}
\end{figure}

@@

While a single-user edits can always be checked to see
whether they create a cycle and thus the system
multi-user edits may not manifest cycles until after
the edits are merged.
there are much more practical reasons why cycles
may appear when multiple users are editing part of the code.

For example, consider the situation shown in Figure~\ref{fig:Multi-User Cycles}.
The top graph represents code that is synchronized between Alice and Bob
and represents the code \texttt{(1 * \hole) * (2 * \hole)}.
The left graph shows

>>>>

This section introduces collaborative structure editing in Grove by example. 
We begin in Section~\ref{sub:Program Representation} by introducing basic edit actions
and the underlying graph representation of expressions with holes.
We will then discuss scenarios that arise in the collaborative setting in 
Section~\ref{sub:Collaborative Structure Editing}.

@@

TODO: typewritter font in figure graphs

One is the underlying graph model, which
involves explicit identifiers, graph edges, indexes, and add and delete
operations.

We present this at two levels.
The other is the interface a user might see,
which omits these details unless needed.
Throughout these examples the graph model is presented
as a graph with.

TODO: somewhere in here, maybe multiple places, we want to talk about typing the intermediate states that arise (background for structured editing)

\subsection{Program Representation}
\label{sub:Program Representation}

\begin{figure}
\centering
TODO: UI

\includegraphics[scale=0.75]{graphviz/simple-tree-with-hole/1.dot}
\caption{A simple tree with a hole}
\label{fig:A simple tree with a hole}
\end{figure}

The editor state for each replica in Grove consists of an expression with holes.
For example, Figure~TODO:1(a) shows the simple arithmetic expression \texttt{2 + \hole}, which
has a hole on the right side of the plus operator.

Internally, Grove represents an expression like this as a directed graph with a distinguished root vertex.
Each vertex represents a term and is labeled with a globally unique identifier and a term constructor.
In this figure, vertices~3 and~6 have the constructors \texttt{plus} and \texttt{num(2)}, respectively.
Each term constructor has an associated set of child positions.
The \texttt{plus} constructor has positions for \texttt{left} and \texttt{right} children
and the \texttt{num(n)} constructors have no child positions.
The root vertex has a distinguished identifier and a constructor with a single child position.
For simplicity, we visually represent the root with a dot.

Each edge is labeled with a globally unique identifier (e.g.,~4 and~7 in the figure) and a child position.
An edge indicates that the destination vertex is a child of the origin vertex at the given position.
Visually, we indicate the position of an edge by the location of its origin.
For example, Figure~TODO:1(b) shows the graph corresponding to the term in Figure~TODO:1(a).

TODO: holes

% In this paper we represent variables and numeric constants
% as leaves and represent their names or values as
% arguments to those constructors (e.g., \texttt{2} in \texttt{num(2)}).
% (In TODO:REF, we discuss how to allow edits within these.)

\subsection{Single User Edits}
\label{sub:Single User Edits}

TODO: change num(2) in Fig1 to num(1) so Fig1 matches Fig2 (move parts of Fig 2 into Fig 1?)

TODO: remove cursor from Fig 1 and Fig 2

(though as we show in Section~\ref{sec:formalism} creating a vertex is implicit in edge creation)

Continuity of examples: Add -> Change -> Wrap -> Move

\subsubsection{Add}
TODO: explain add to the right

TODO: continuity of examples

\subsubsection{Change}

@@

TODO: once deleted an edge can never be recreated (though, a new edge (with a new id) can be created in the same spot as the original)

@@

\footnote{For the purposes of this paper, we represent identifiers and number literals
indivisibly.  See Section~TODO:REF, for how we would allow editing individual characters.}

TODO: 2 -> 22

@@

\subsection{Collaborative Structure Editing}
\label{sub:Collaborative Structure Editing}

@@

\begin{figure}
\centering
TODO: lam x. app(\#1, \hole)
\caption{Single-User Cycles}
\label{fig:Single-User Cycles}
\end{figure}

@@

While a single-user edits can always be checked to see
whether they create a cycle and thus the system
can immediately warn the user,
multi-user edits may not manifest cycles until after
the edits are merged.

For example, consider the situation shown in Figure~\ref{fig:Multi-User Cycles}.
The top graph represents code that is synchronized between Alice and Bob
and represents the code \texttt{(1 * \hole) * (2 * \hole)}.
In the left graph, Alice moves vertex~TODO to the \texttt{right}
index of vertex~TODO.
The resulting graph represents the code \texttt{(1 * (2 * \hole)) * \hole}.
In the right graph, Bob moves vertex~TODO to the \texttt{right} index of vertex~TODO.
The resulting graph represents the code \texttt{\hole * (2 * (1 * \hole))}.
On their own, neither edit creates a cycle.
However, when these edits are merged,
not only is a cycle created, but
the vertexes in the cycle

Can also be detected due to neigther original leaving them disconnected.
